### 2014-04-07 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding ML results to multilevel tech note

\begin{abstract}
  Monte Carlo solvers for linear systems have been demonstrated to
  perform poorly for strongly elliptic problems. This poor performance
  is primarily due to the fact that both a large number of samples are
  required to obtain a good statistical error and that these samples
  require a large amount of time to compute. As a mechanism to reduce
  the computational complexity for such problems and thereby improve
  the figure of merit of the calculation, multilevel Monte Carlo has
  been introduced in finance problems, the solution of stochastic
  partial differential equations, and other algorithms that leverage
  Markov chain Monte Carlo. We adapt these ideas to form a multilevel
  Monte Carlo solver for linear systems which, in the form presented
  here, is effectively a stochastic realization of a multigrid
  solver. Numerical studies indicate that the new multilevel method
  can reduce the time required to achieve a certain statistical error
  in the solution by at least two orders of magnitude, thereby
  dramatically reducing the computational complexity of the
  problem. Furthermore, the general formulation presented here
  indicates that this methodology is not restricted to geometric
  formulations of the multigrid method and could additionally be
  formulated using algebraic techniques.
\end{abstract}

@@

  \end{algorithmic}
\end{algorithm}
This algorithm permits each of the levels to be computed independently
and then combined appropriately with the prolongation and restriction
operators. Like other multigrid algorithms for linear systems, we are
required to produce a representation of the linear operator on each
level, $\ve{A}_l$, as well as the right hand side, $\ve{b}_l$. The
formulation of Algorithm~\ref{alg:mlamc} places no restrictions on the
form of $\ve{A}_l$, $\ve{b}_l$, $\ve{P}_l$, or $\ve{R}_l$
and therefore could be used with any multigrid scheme including those
arising from algebraic multigrid formulations.

@@

We again perform calculations with $N=10,000$ and initial guesses of
wave numbers 1, 5 and 10 and vary the number of levels used in the
calculation. Tables~\ref{tab:k1_results}, \ref{tab:k5_results}, and
\ref{tab:k10_results} giving the results of these calculations. In
these tables, the RFOM column indicates the relative figure of merit
which has been normalized to the single level calculation. The results
indicate the performance of the multilevel method compared to the
original adjoint Monte Carlo method is excellent, offering a dramatic
improvement in time to solution and figure of merit values 10-100
times that of the single level solver.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 0.022 & 119.1 & 1 \\
      2 & 13,535 & 0.026 & 72.0 & 1.14 \\
      3 & 14,785 & 0.035 & 33.2 & 1.38 \\
      4 & 15,226 & 0.039 & 13.7 & 2.63 \\
      5 & 15,382 & 0.027 & 5.5 & 13.86 \\
      6 & 15,437 & 0.036 & 2.3 & 19.21 \\
      7 & 15,456 & 0.045 & 0.98 & 28.00 \\
      8 & 15,462 & 0.081 & 0.41 & 20.66 \\
      9 & 15,464 & 0.267 & 0.17 & 11.72 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Multilevel Monte Carlo results for $k = 1$ with $M
      = 2$, $N = 10,000$ and run times reported in seconds.}}
  \label{tab:k1_results}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 0.668 & 101.6 & 1 \\
      2 & 13,535 & 0.381 & 60.0 & 5.21 \\
      3 & 14,785 & 0.396 & 28.0 & 10.34 \\
      4 & 15,226 & 0.599 & 11.9 & 10.64 \\
      5 & 15,382 & 0.638 & 4.7 & 23.60 \\
      6 & 15,437 & 1.052 & 1.8 & 23.15 \\
      7 & 15,456 & 1.070 & 0.67 & 59.16 \\
      8 & 15,462 & 1.180 & 0.23 & 144.10 \\
      9 & 15,464 & 2.130 & 0.10 & 99.95 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Multilevel Monte Carlo results for $k = 5$ with $M
      = 2$, $N = 10,000$ and run times reported in seconds.}}
  \label{tab:k5_results}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 1.81 & 98.6 & 1 \\
      2 & 13,535 & 1.67 & 59.6 & 1.94 \\
      3 & 14,785 & 2.41 & 27.1 & 2.05 \\
      4 & 15,226 & 3.20 & 11.5 & 2.74 \\
      5 & 15,382 & 1.85 & 4.93 & 19.14 \\
      6 & 15,437 & 3.89 & 1.95 & 10.95 \\
      7 & 15,456 & 3.79 & 0.78 & 28.68 \\
      8 & 15,462 & 7.08 & 0.30 & 21.62 \\
      9 & 15,464 & 11.7 & 0.11 & 21.45 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
      = 2$, $N = 10,000$ and run times reported in seconds.}}
  \label{tab:k10_results}
\end{table}

@@

  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
      = 4$, $N = 10,000$ and run times reported in seconds.}}

@@

  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
      = 4$, $N = 100,000$ and run times reported in seconds.}}

@@

  \caption{\textbf{Multilevel Monte Carlo results for $k = 10$ with $M
      = 4$, $N = 1,000,000$ and run times reported in seconds.}}

@@

%%---------------------------------------------------------------------------%%
\section{Algebraic Multigrid Results}
\label{sec:algebraic_results}

>>>>

\begin{abstract}
  Monte Carlo solvers for linear systems have been demonstrated to
  perform poorly for strongly elliptic problems. This poor performance
  is primarily due to the fact that both a large number of samples are
  required to obtain a good statistical error and that these samples
  require a large amount of time to compute. As a mechanism to reduce
  the computational complexity for such problems and thereby improve
  the figure of merit of the calculation, multilevel Monte Carlo has
  been introduced in finance problems, the solution of stochastic
  partial differential equations, and other algorithms that leverage
  Markov chain Monte Carlo. We adapt these ideas to form a multilevel
  Monte Carlo solver for linear systems which, in the form presented
  here, is effectively a stochastic realization of a multigrid
  solver. Numerical studies indicate that the new multilevel method
  can reduce the time required to achieve a certain statistical error
  in the solution by at least two orders of magnitude, thereby
  dramatically reducing the computational complexity of the
  problem. Furthermore, the general formulation and results presented
  here indicate that this methodology may be used with both geometric
  and algebraic formulations of the Galerkin mutlgrid method.
\end{abstract}

@@

  \end{algorithmic}
\end{algorithm}
This algorithm permits each of the levels to be computed independently
and then combined appropriately with the prolongation and restriction
operators. Like other multigrid algorithms for linear systems, we are
required to produce a representation of the linear operator on each
level, $\ve{A}_l$, as well as the right hand side, $\ve{b}_l$. The
formulation of Algorithm~\ref{alg:mlamc} places no restrictions on the
form of $\ve{A}_l$, $\ve{b}_l$, $\ve{P}_l$, or $\ve{R}_l$
and therefore could be used with any multigrid scheme including those
arising from algebraic multigrid formulations. It should be noted,
however, that work to date has only tested the algorithm for the
Galerkin formulation presented in Algorithm~\ref{alg:mlamc}.

@@

We again perform calculations with $N=10,000$ and initial guesses of
wave numbers 1, 5 and 10 and vary the number of levels used in the
calculation. Tables~\ref{tab:k1_results}, \ref{tab:k5_results}, and
\ref{tab:k10_results} giving the results of these calculations. In
these tables, the RFOM column indicates the relative figure of merit
which has been normalized to the single level calculation. The results
indicate the performance of the multilevel method compared to the
original adjoint Monte Carlo method is excellent, offering a dramatic
improvement in time to solution and figure of merit values 10-100
times that of the single level solver.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 0.022 & 119.1 & 1 \\
      2 & 13,535 & 0.026 & 72.0 & 1.14 \\
      3 & 14,785 & 0.035 & 33.2 & 1.38 \\
      4 & 15,226 & 0.039 & 13.7 & 2.63 \\
      5 & 15,382 & 0.027 & 5.5 & 13.86 \\
      6 & 15,437 & 0.036 & 2.3 & 19.21 \\
      7 & 15,456 & 0.045 & 0.98 & 28.00 \\
      8 & 15,462 & 0.081 & 0.41 & 20.66 \\
      9 & 15,464 & 0.267 & 0.17 & 11.72 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
      1$ with $M = 2$, $N = 10,000$ and run times reported in
      seconds.}}
  \label{tab:k1_results}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 0.668 & 101.6 & 1 \\
      2 & 13,535 & 0.381 & 60.0 & 5.21 \\
      3 & 14,785 & 0.396 & 28.0 & 10.34 \\
      4 & 15,226 & 0.599 & 11.9 & 10.64 \\
      5 & 15,382 & 0.638 & 4.7 & 23.60 \\
      6 & 15,437 & 1.052 & 1.8 & 23.15 \\
      7 & 15,456 & 1.070 & 0.67 & 59.16 \\
      8 & 15,462 & 1.180 & 0.23 & 144.10 \\
      9 & 15,464 & 2.130 & 0.10 & 99.95 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
      5$ with $M = 2$, $N = 10,000$ and run times reported in
      seconds.}}
  \label{tab:k5_results}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 1.81 & 98.6 & 1 \\
      2 & 13,535 & 1.67 & 59.6 & 1.94 \\
      3 & 14,785 & 2.41 & 27.1 & 2.05 \\
      4 & 15,226 & 3.20 & 11.5 & 2.74 \\
      5 & 15,382 & 1.85 & 4.93 & 19.14 \\
      6 & 15,437 & 3.89 & 1.95 & 10.95 \\
      7 & 15,456 & 3.79 & 0.78 & 28.68 \\
      8 & 15,462 & 7.08 & 0.30 & 21.62 \\
      9 & 15,464 & 11.7 & 0.11 & 21.45 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
      10$ with $M = 2$, $N = 10,000$ and run times reported in
      seconds.}}
  \label{tab:k10_results}
\end{table}

@@

  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
      10$ with $M = 4$, $N = 10,000$ and run times reported in
      seconds.}}

@@

  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
      10$ with $M = 4$, $N = 100,000$ and run times reported in
      seconds.}}

@@

  \caption{\textbf{Geometric Multilevel Monte Carlo results for $k =
      10$ with $M = 4$, $N = 1,000,000$ and run times reported in
      seconds.}}

@@

%%---------------------------------------------------------------------------%%
\section{Algebraic Multigrid Results}
\label{sec:algebraic_results}
We next solve the same model problem, this time using an algebraic
multigrid formulation of the Galerkin problem. For these calculations
we used the ML multigrid library to construct $\ve{A}$, $\ve{P}$, and
$\ve{R}$ on each level with $\ve{M} \approx 3$
\cite{ml_manual}. Tables~\ref{tab:ml_k1_results},
\ref{tab:ml_k5_results}, and \ref{tab:ml_k10_results} give the results
of these calculations.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 0.022 & 119.1 & 1 \\
      2 & 11,924 & 0.037 & 31.9 & 1.28 \\
      3 & 12,294 & 0.055 & 7.6 & 2.40 \\
      4 & 12,365 & 0.051 & 1.8 & 11.76 \\
      5 & 12,378 & 0.094 & 0.5 & 12.01 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Algebraic Multilevel Monte Carlo results for $k =
      1$ with $M = 3$, $N = 10,000$ and run times reported in
      seconds.}}
  \label{tab:ml_k1_results}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 0.668 & 101.6 & 1 \\
      2 & 11,924 & 0.834 & 30.8 & 2.12 \\
      3 & 12,294 & 1.140 & 7.5 & 4.63 \\
      4 & 12,365 & 0.975 & 1.6 & 29.09 \\
      5 & 12,378 & 2.240 & 0.4 & 25.10 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Algebraic Multilevel Monte Carlo results for $k =
      5$ with $M = 3$, $N = 10,000$ and run times reported in
      seconds.}}
  \label{tab:ml_k5_results}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cllll}\hline\hline
      \multicolumn{1}{c}{\textbf{Levels}} & 
      \multicolumn{1}{l}{\textbf{Samples}} & 
      \multicolumn{1}{l}{\textbf{$||\ve{e}||_{\infty}$}} & 
      \multicolumn{1}{l}{\textbf{Time (s)}} & 
      \multicolumn{1}{l}{\textbf{RFOM}} \\
      \hline
      1 & 10,000 & 1.81 & 98.6 & 1 \\
      2 & 11,924 & 2.33 & 30.0 & 1.99 \\
      3 & 12,294 & 3.19 & 7.6 & 4.20 \\
      4 & 12,365 & 3.31 & 1.8 & 16.29 \\
      5 & 12,378 & 9.98 & 0.5 & 7.21 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Algebraic Multilevel Monte Carlo results for $k =
      10$ with $M = 3$, $N = 10,000$ and run times reported in
      seconds.}}
  \label{tab:ml_k10_results}
\end{table}
In general, the results show similar performance to the geometric
multigrid method, however, gains in figure of merit were not as
significant which may be in part due to the simplified selection of
sample size at each level. In addition, the Dirichlet conditions for
the model problem were explicitly included in the geometric
interpolation operators while the level operators generated by the
algebraic scheme did not. Most importantly, these results indicate
that Algorithm~\ref{alg:mlamc} can be used with general Galerkin
formulations of the multigrid problem.
